---
title: "individual2"
author: "Cody Nielsen"
date: "9/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring General Social Survey

Load data
```{r}
library(tidyverse)
dat <- readxl::read_xls('GSS.xls')
dat
```

Clean data  
```{r}
dat_clean <- dat %>% rename(marital_status='Marital status', years_school='Highest year of school completed', sex='Respondents sex', pol_party='Political party affiliation', happy_lvl='General happiness')
names(dat_clean)
dat_clean
dat_clean <- transform(dat_clean, years_school = as.numeric(years_school))
class(dat_clean$years_school)
```

1.Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.  

```{r}
dat1 <-aggregate(years_school~pol_party, dat_clean, mean)
dat1<- dat1[order(dat1$years_school,decreasing=T),]
dat1
```
  
2.Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

```{r}
#dat2 <-aggregate(years_school~pol_party, dat_clean, median)
#dat2 <- dat_clean %>% filter (years_school>median(years_school))
#str(dat2)
dat2 <-dat_clean %>% 
  select(years_school, pol_party) %>%
  group_by(pol_party) %>% 
  filter(years_school > median(years_school, na.rm = TRUE))
str(dat2)
```
  
3.How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.  

```{r}
dat_clean$happy_lvl <- sub('Very happy', 'Happy', dat_clean$happy_lvl)
dat_clean$happy_lvl <- sub('Pretty happy', 'Happy', dat_clean$happy_lvl)
dat_clean<-na.omit(dat_clean)
ggplot(dat_clean, aes(x=marital_status, fill=happy_lvl))+geom_bar()+xlab("Marital status")
```  
  
Looks like people are more happy married.  

4.Does party affiliation affect the relationship you found in the last question?  
```{r}
dat3 <-dat_clean %>% select(marital_status,happy_lvl, pol_party)
g <- ggplot(data = dat3, aes(x = marital_status, fill = happy_lvl)) + geom_bar(position = "fill") +coord_flip()+facet_wrap(~pol_party)
g
```  

Doesn't look like political party affiliation has any effect, people seem generally happy.  

5.How is marital status related to the education of a respondent? What are possible reasons accounting for the patterns you see? 

```{r}
ggplot(dat_clean, aes(years_school)) + facet_wrap(~marital_status)+geom_histogram(bins=10)
```  

It looks the highest frequency of educated people are married. The frequency of education is lower for separated couples. This could be because educated couples make better decisions.  

6.Explore two more interesting questions, and answer those questions using the GSS data.

A) Does sex affect political party affiliation?  

```{r}
ggplot(dat_clean, aes(x=pol_party, fill=sex))+geom_bar()+coord_flip()+theme_bw()+facet_wrap(~sex, scales="free_x")

ggplot(dat_clean, aes(x=pol_party,group=sex,fill=sex))+geom_density(adjust=1.5, alpha=.4)+theme(axis.text.x = element_text(angle = 90))
```  

B) What is the relationship between education and sex?  

```{r}
g1 <- ggplot(dat_clean, aes(x=years_school, fill=sex))+geom_bar()+facet_wrap(~sex, scales="free_x")
g1 + scale_fill_brewer(palette="Dark2")

g2 <- ggplot(dat_clean, aes(x=years_school, fill=sex))+geom_density()+facet_wrap(~sex, scales = "free_x")
g2

g3 <- ggplot(dat_clean, aes(x=years_school, group=sex, fill=sex))+geom_density(adjust=1.5, alpha=.4)
g3
```